---
title: "BMS353 Week 6"
author: "Mark Dunning"
date: "November 2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggthemr)
uos_pal <- function(){
  uos_pal <- c("Process_Cyan"=rgb(0,159,218,maxColorValue = 255),
               "Pantone_274"=rgb(31,20,93,maxColorValue = 255),
               "Process_Yellow"=rgb(249,227,0,maxColorValue = 255),
               
               "Pantone_347"=rgb(0,155,72,maxColorValue = 255),
               "Pantone_382"=rgb(190,214,0,maxColorValue = 255),
               "Process_Magenta"=rgb(209,0,116,maxColorValue = 255),
               "Pantone_Orange_021"=rgb(255,88,0,maxColorValue = 255),
               
               "Pantone_512"=rgb(119,33,111,maxColorValue = 255),
               "Pantone_485"=rgb(213,43,30,maxColorValue = 255),
               "Pantone_Black"=rgb(30,30,30,maxColorValue = 255),
               "Pantone_161"=rgb(98,60,27,maxColorValue = 255),
               
               "Pantone_7501"=rgb(219,206,172,maxColorValue = 255),
               "Pantone_343"=rgb(3,86,66,maxColorValue = 255),
               "Pantone_322"=rgb(0,116,122,maxColorValue = 255),
               "Pantone_202"=rgb(130,36,51,maxColorValue = 255)
  )
  
  uos_pal
}

uos_colours <- as.character(uos_pal())
# you have to add a colour at the start of your palette for outlining boxes, we'll use a grey:
uos_colours <- c("#555555", uos_colours)
# remove previous effects:

ggthemr_reset()
# Define colours for your figures with define_palette
uos <- define_palette(
  swatch = uos_colours, # colours for plotting points and bars
  gradient = c(lower = uos_colours[1L], upper = uos_colours[2L]), #upper and lower colours for continuous colours
  background = "white" #defining a grey-ish background 
)
# set the theme for your figures:
ggthemr(uos)
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(DESeq2))
suppressPackageStartupMessages(library(ggrepel))
dds <- makeExampleDESeqDataSet()
```

## Week 5 Recap

- Adding extra annotation to differential expression results
  - Bioconductor packages
  - biomaRt
- Visualisations to help interogate results
  - heatmaps
  - volcano plots



## Week 6 Overview

- Towards biological interpretation
- What pathways or biological processes explain the gene list
- Two types of test
  - Apply significance cut-off to the results
  - Rank the gene list
- Overview of assignment

## So you've got your gene list: what now?

- We have used an annotation resource to convert our IDs to something more meanginful
- Useful if we have a particular positive of negative control
  + e.g. a gene that has been knocked-out
  + genes know from the literature

```{r}
library(dplyr)
results_annotated <- readRDS("../Robjects/results_cellType_annotated.rds")
results_annotated %>% 
  dplyr::select(ENSEMBL,padj,SYMBOL) %>% 
  head(5)
```

## So you've got your gene list: what now?

- The same resource can be used to obtain membership of *gene ontologies*
  + collection of genes that share biological function
- There are lots of pathways identified, so we need a statistical test to determine if any occur more often than we expect.

```{r message=FALSE}
library(org.Mm.eg.db)
anno <- AnnotationDbi::select(org.Mm.eg.db, keys = results_annotated$ENSEMBL, keytype = "ENSEMBL",columns="GO")
results_annotated %>% 
  left_join(anno) %>% 
  dplyr::select(ENSEMBL,padj,SYMBOL,GO) %>% 
  group_by(ENSEMBL) %>% 
  mutate(GO = paste(GO,collapse=",")) %>% 
  dplyr::select(-ENSEMBL) %>% 
  distinct %>% 
  head(5)

```

## Testing for Significance

- We can count up how many times each pathway is seen
- But this can be deceiving
  + not all pathways are of equal size (from a few genes to a few thousand)
- If the gene list were random, with no biological significance, we would still get a particular distribution of pathways


```{r}
de_genes <- results_annotated %>% 
  filter(padj < 0.05, abs(log2FoldChange) > 1)
de_GO <- AnnotationDbi::select(org.Mm.eg.db, keys = de_genes$ENSEMBL,
                      keytype = "ENSEMBL",
                      columns = "GO")
dplyr::count(de_GO, GO) %>% 
  arrange(desc(n))
```
## Testing for Significance

- Aim is to compare number of differentially-expressed genes for each pathway - to how many we expect by chance
- This can be done using a *Fisher's exact test* or chi-squared test (R will take care of this for us)
  + $$ p = \frac{\binom{a + b}{a}\binom{c +d}{c}}{\binom{n}{a +c}} = \frac{(a+b)!(c+d)!(a+c)!(b+d)!}{a!b!c!d!n!} $$
- Result is p-value for *each pathway*

```{r}
df <- data.frame(`In DE List`  = c("a","c","a+c"), `Not in DE list` = c("b","d","b+d"), RowTotal = c("a +b","c+d","a+b+c+d (=n)"))
rownames(df) <- c("In Gene Set", "Not in Gene Set","Column Total")
df
```


## Alternative approach

- All of the above assumes we impose a threshold on our data
- Are genes with p-value 0.051 any less significant? Or genes with log-fold change 0.99?
- What happens if we don't get any, or very few, genes?
- An alternative approach exists that takes information from *all genes*

## The GSEA method

- The Gene Set Enrichment Analysis (GSEA) tool created by the Broad Institute can be used in such circumstances
- No threshold are required, but the results must be ranked by a criteria with directionality such as log fold-change
- Seeks to identify pathways with a tendency to be at top or bottom of ranked gene list

## The GSEA method

- For each gene set in turn, walk-down the gene list and record every time a gene belonging to that pathway is seen
- For a *down-regulated* gene set, the genes belonging to it would be concentrated towards the *bottom* of the gene list

```{r}
suppressPackageStartupMessages(library(limma))
stat <- sort(rnorm(100))
sel <- c(35,50,sample(1:20, 5))
barcodeplot(stat, index = sel, main="Down-regulated Gene Set")

```

## The GSEA method

- For an *up-regulated* gene set, the genes belonging to it would be concentrated towards the *top* of the gene list

```{r}
sel2 <- c(45,60, sample(80:100,7))
barcodeplot(stat, index = sel2, main="Up-regulated Gene Set")
```

## The GSEA method



```{r}
sel3 <- sample(1:100, 9)
barcodeplot(stat, index = sel3, main="Neither up- or down-regulated")
```


## Assignment Overview

## Further Viewing

